---
title: "Barcode validation analysis"
author: '@rvosa'
date: "2025-02-23"
output: pdf_document
---

# BGE Project Milestone MS8.4.1: DNA Barcode Validation Analysis

This document analyzes DNA barcode validation results from the BGE project, 
combining outputs from two institutions (Naturalis and Natural History Museum) 
with BOLD spreadsheet data. The analysis explores parameter optimization, 
sequence quality metrics, temporal patterns, and taxonomic success rates. The 
results will inform optimization of the MGE pipeline parameters and contribute 
to understanding factors affecting DNA barcode recovery from historical 
specimens.

```{r}
source("src/functions/data_loading.R")

validation_data <- load_validation_data(
  nhm_path = "data/concatenated_untriaged.tsv",
  naturalis_path = "data/sum.tsv"
)

# Access individual dataframes
nhm_df <- validation_data$nhm
naturalis_df <- validation_data$naturalis
```

```{r}
source("src/functions/preprocessing.R")

joined_data <- join_validation_taxonomy(
  validation_data,
  lab_sheet_path = "data/lab_sheet.tsv",
  taxonomy_path = "data/taxonomy.tsv"
)
```